Output Summary

### 2021-08-03 03-post-process/output-plots.Rmd
Completely revamp MCMC diagnostics in output-plots
* All info is in a table now, no MCMC diagnostic plots. This compiles much faster and shows all information more compactly
* Information is shown (% of nodes that violate a threshold) for Rhat, ESS (mean & quantiles), and MCMC CV (mean & quantiles)
* All estimated parameters are now diagnosed, where as before some were missing
* Diagnostic summaries are calculated using the updated calculations included in the 'posterior' package
* Table is color-coded based on severity of violations

This table shows the number of parameters that fail to meet some diagnostic threshold. Lower Rhat values are good, larger Neff values are good, and lower MCMC SE/Mean values are good.

The "match" column shows the kinds of parameters summarized in that row. E.g., `^mu_phi` covers all parameter names that start with `mu_phi` and are followed by anything else (`r paste(match_params(post, "^mu_phi"), collapse = ", ")`).

This model was ran for a very short time, and I'm impressed with how well the sampling performed even considering this.

```{r mcmc-diags}
# groups of parameters to match and calculate diagnostic summaries for separately
diag_params = list(
  c("alpha", "beta", "sigma_Pb"),
  "gamma",
  "^mu_phi",
  "^mu_psi",
  "^mu_omega",
  "^mu_pi",
  "^sig_Lphi",
  "^sig_Lpsi",
  c("^sig_Lomega", "^sig_Lpi"),
  "^Pb[", 
  "init_recruits", 
  "^phi_Ma_O0", "^phi_Mb_Ma[.+,2]", "^phi_Pa_Mb[.+,1]", "^phi_Pa_Mb[.+,2]",
  "^phi_O0_O1", "^phi_O1_O2", "^phi_O2_O3",
  "^psi_O1_Rb", "^psi_O2_Rb", "O_phi_scaler_nat_hat"
)

# calculate the diagnostic statistics for each group
diags = lapply(diag_params, function(param) as.matrix(post_summ(post, param, Rhat = T, neff = T)[c("mean", "Rhat", "neff"),]))

# count up the number of parameters by group that fail to meet certain thresholds
diag_summ = t(sapply(diags, function(param) {
  c(
    n_total = ncol(param),
    n_Rhat_1 = sum(param["Rhat",] > 1.05),
    n_Rhat_2 = sum(param["Rhat",] > 1.1),
    n_Rhat_3 = sum(param["Rhat",] > 1.2),
    n_neff_1 = sum(param["neff",] < 2000),
    n_neff_2 = sum(param["neff",] < 1000),
    n_neff_3 = sum(param["neff",] < 500)
  )
}))

# format to a table
diag_summ = as.data.frame(diag_summ)
colnames(diag_summ) = c("Total Params", ">1.05", ">1.1", ">1.2", "<3000", "<1000", "<500")
df = cbind("Match" = unlist(lapply(diag_params, paste, collapse = ", ")), diag_summ)

# render the table
kable(df, align = "lccccccc") %>%
  kable_styling(full_width = F, bootstrap_options = c("condensed", "striped")) %>%
  add_header_above(c(" " = 2, "Rhat" = 3, "Neff" = 3)) %>%
  column_spec(1, monospace = T) %>%
  column_spec(c(2,5), border_right = T)

@@

`r if (params$mcmc_plots) "## Diagnostic Plots {.tabset .tabset-pills}"`

`r if (params$mcmc_plots) "### $\\alpha$, $\\beta$"`

```{r alpha-beta-diag-plots, eval = params$mcmc_plots, fig.width = 4, fig.height = 5}
diag_plots(post, c("alpha", "beta"), layout = "2x1")
```

`r if (params$mcmc_plots) "### $\\gamma$"`

```{r gamma-diag-plots, eval = params$mcmc_plots, fig.width = 3, fig.height = 6}
diag_plots(post, "gamma", layout = "4x1")
```

`r if (params$mcmc_plots) "### $\\mu$"`

```{r mu-diag-plots, eval = params$mcmc_plots, fig.width = 7, fig.height = 7}
diag_plots(post, "^mu", layout = "5x3")
```

`r if (params$mcmc_plots) "### $\\sigma$"`

```{r sig-diag-plots, eval = params$mcmc_plots, fig.width = 7, fig.height = 7}
diag_plots(post, "^sig", layout = "5x3")
```

`r if (params$mcmc_plots) "### init_recruits"`

```{r init-recruits-diag-plots, eval = params$mcmc_plots, fig.width = 5, fig.height = 7}
diag_plots(post, "init_recruits", layout = "4x2")
```

`r if (params$mcmc_plots) "### surv_scaler"`

```{r surv-scaler-diag-plots, eval = params$mcmc_plots, fig.width = 6, fig.height = 4}
diag_plots(post, "O_phi_scaler_nat_hat")
```

`r if (params$mcmc_plots) "### stray"`

_There are some elements of both `n_stray_tot` and `stray_comp` that are exactly zero each MCMC iteration. These are omitted from the plots below._

```{r stray-diag-plots, eval = params$mcmc_plots, fig.width = 7, fig.height = 7}
stray = post_subset(post, "stray", T)
non_zero_stray = apply(stray, 2, function(x) all(x != 0))
stray_nodes = names(non_zero_stray)[non_zero_stray]
diag_plots(post, stray_nodes, layout = "5x3")
```

>>>>

```{r mcmc-diag-function}
# a function to get MCMC diagnoses for a group of parameters
# params: a string to match with postpack::match_params()
# L1: a "Level-1" group label; such as component of the model (e.g., LH apportionment)
# L2: a "Level-2" group label; such as type of parameter (e.g., mean, sd, or year-specific value)
# L3: a "Level-3" group label; such as NOR or females or age-3 females

diag_fn = function(params, L1 = NA, L2 = NA, L3 = NA) {

  # extract the posteriors of interest and format for 'posterior' package
  post_sub = as_draws_df(post_subset(post, params))

  # summarize the posteriors
  diags = summarize_draws(post_sub, mean, ~quantile(.x, c(0.025, 0.975)), rhat, ess_mean, ~ess_quantile(.x, probs = c(0.025, 0.975)), mcse_mean, ~mcse_quantile(.x, probs = c(0.025, 0.975)))

  # discard any parameters that were the same value every MCMC iteration
  # (will have an NA for rhat)
  diags = diags[!is.na(diags$rhat),]

  # add "Monte Carlo CV"
  diags$mccv_mean = diags$mcse_mean/diags$mean
  diags$mccv_q2.5 = diags$mcse_q2.5/diags$`2.5%`
  diags$mccv_q97.5 = diags$mcse_q97.5/diags$`97.5%`

  # summarize the summaries
  out = data.frame(
    L1 = L1,
    L2 = L2,
    L3 = L3,
    match = params,
    n_params = nrow(diags),
    Rhat_1.05 = sum(diags$rhat >= 1.05),
    Rhat_1.1 = sum(diags$rhat >= 1.1),
    Rhat_1.2 = sum(diags$rhat >= 1.2),
    ess_mean_400 = sum(diags$ess_mean < 400),
    ess_q2.5_400 = sum(diags$ess_q2.5 < 400),
    ess_q97.5_400 = sum(diags$ess_q97.5 < 400),
    mccv_mean_0.05 = sum(diags$mccv_mean > 0.05),
    mccv_q2.5_0.05 = sum(diags$mccv_q2.5 > 0.05),
    mccv_q97.5_0.05 = sum(diags$mccv_q97.5 > 0.05)
  )

  # return the summarized summaries
  return(out)
}

@@

```{r calculate-mcmc-diagnostics}
# parr recruitment
L1 = "Parr Recruitment"
L2 = c("BH Productivity", "BH Capacity", "Process SD", "Parr States")
match = c("alpha", "^beta[", "sigma_Pb", "^Pb[")
parr_recruit_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# LH apportionment
L1 = "Life History Apportionment"
L2 = c("Mean", "Process SD", "Realized")
match = c("mu_pi[1,.]", "sig_Lpi", "^pi[.+,1,.]")
pi_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Overwinter survival
L1 = "Overwinter Survival"
L2 = c("Intercept", "Slope", "Process SD", "Realized")
base = list("gamma0[LH_type,pop]", "gamma1[LH_type,pop]", "sig_Lphi_Pa_Mb[LH_type,pop]", "^phi_Pa_Mb[year,LH_type,pop]")
match = sub_index(base, year = ".+", LH_type = i_fall, pop = ".")
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Fall Migrants")))
match = sub_index(base, year = ".+", LH_type = i_spring, pop = ".")
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Spring Migrants")))
phi_Pa_Mb_diags = rbind(tmp1, tmp2)

# Migration survival: trib to LGR
L1 = "Migration Survival (Trib to LGR)"
L2 = c("Mean", "Process SD", "Realized")
base = list("mu_phi_Mb_Ma[LH_type,origin,pop]", "sig_Lphi_Mb_Ma[LH_type,origin,pop]", "phi_Mb_Ma[year,LH_type,origin,pop]")
match = sub_index(base, year = ".+", LH_type = i_spring, origin = o_nor, pop = ".")
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "NOR")))
match = sub_index(base, year = ".+", LH_type = i_spring, origin = o_hor, pop = ".")
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "HOR")))
phi_Mb_Ma_diags = rbind(tmp1, tmp2)

# Migration survival: LGR to BON
L1 = "Migration Survival (LGR to Ocean)"
L2 = c("Mean", "Process SD", "Realized")
base = c("mu_phi_Ma_O0[origin]", "sig_Lphi_Ma_O0[origin]", "phi_Ma_O0[year,origin]")
match = sub_index(base, year = ".+", origin = o_nor)
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "NOR")))
match = sub_index(base, year = ".+", origin = o_hor)
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "HOR")))
phi_Ma_O0_diags = rbind(tmp1, tmp2)

# Sex Apportionment
L1 = "Sex Apportionment"
L2 = c("Mean", "Process SD", "Realized")
base = list("mu_omega[sex,origin,pop]", "sig_Lomega[origin,pop]", "omega[year,sex,origin,pop]")
match = sub_index(base, year = ".+", sex = s_female, origin = o_nor, pop = ".")
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "NOR")))
match = sub_index(base, year = ".+", sex = s_female, origin = o_hor, pop = ".")
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "HOR")))
omega_diags = rbind(tmp1, tmp2)

# Probability of Maturing at Age-3
L1 = "Pr(Mature at Age-3)"
L2 = c("Mean", "Process SD", "Realized")
base = c("mu_psi_O1_Rb[sex,origin,pop]", "sig_Lpsi_O1_Rb[sex,origin,pop]", "psi_O1_Rb[.+,sex,origin,pop]")
match = sub_index(base, year = ".+", sex = s_female, origin = o_nor, pop = ".")
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Female (NOR)")))
match = sub_index(base, year = ".+", sex = s_female, origin = o_hor, pop = ".")
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Female (HOR)")))
match = sub_index(base, year = ".+", sex = s_male, origin = o_nor, pop = ".")
tmp3 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Male (NOR)")))
match = sub_index(base, year = ".+", sex = s_male, origin = o_hor, pop = ".")
tmp4 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Male (HOR)")))
psi_O1_Rb_diags = rbind(tmp1, tmp2, tmp3, tmp4)

# Probability of Maturing at Age-4
L1 = "Pr(Mature at Age-4)"
L2 = c("Mean", "Process SD", "Realized")
base = c("mu_psi_O2_Rb[sex,origin,pop]", "sig_Lpsi_O2_Rb[sex,origin,pop]", "psi_O2_Rb[.+,sex,origin,pop]")
match = sub_index(base, year = ".+", sex = s_female, origin = o_nor, pop = ".")
tmp1 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Female (NOR)")))
match = sub_index(base, year = ".+", sex = s_female, origin = o_hor, pop = ".")
tmp2 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Female (HOR)")))
match = sub_index(base, year = ".+", sex = s_male, origin = o_nor, pop = ".")
tmp3 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Male (NOR)")))
match = sub_index(base, year = ".+", sex = s_male, origin = o_hor, pop = ".")
tmp4 = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i], L3 = "Male (HOR)")))
psi_O2_Rb_diags = rbind(tmp1, tmp2, tmp3, tmp4)

# Year 1 ocean survival
L1 = "Year 1 Ocean Survival"
L2 = c("Mean", "Process SD", "Realized")
base = c("mu_phi_O0_O1[origin]", "sig_Lphi_O0_O1", "phi_O0_O1[year,origin]")
match = sub_index(base, year = ".+", origin = o_nor)
phi_O0_O1_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Year 2 ocean survival
L1 = "Year 2 Ocean Survival"
L2 = c("Mean", "Process SD", "Realized")
base = c("mu_phi_O1_O2[origin]", "sig_Lphi_O1_O2", "phi_O1_O2[year,origin]")
match = sub_index(base, year = ".+", origin = o_nor)
phi_O1_O2_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Misc ocean survival
L1 = "Misc Ocean Survival"
L2 = c("NOR:HOR Scaler", "Year 0 Residual", "AR(1) Coefficient")
match = c("O_phi_scaler_nat_hat", "Lphi_O0_O1_resid[5]", "kappa_phi_O0_O1")
misc_ocean_surv_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Pre-spawn Survival
L1 = "Pre-spawn Survival"
L2 = c("Mean", "Process SD", "Realized")
match = c("mu_phi_Sb_Sa[.]", "sig_Lphi_Sb_Sa[.]", "phi_Sb_Sa[.+,.]")
phi_Sb_Sa_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Carcass Correction
L1 = "Carcass Correction Factor"
z_diags = diag_fn("z", L1 = L1)

# Initial adult recruitment
L1 = "Initial Adult Recruitment"
L2 = c("Mean", "Process SD", "Realized")
match = c("mu_init_recruits[.]", "sig_init_lrecruits[.]", "^init_recruits[.+,.]")
init_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# Straying
L1 = "Strays"
L2 = c("Realized Number", "Age/Sex Comp")
match = c("n_stray_tot[.+,.]", sub_index("stray_comp[age,sex,origin,pop]", age = ".", sex = ".", origin = o_hor, pop = "."))
stray_diags = unlist_dfs(lapply(1:length(match), function(i) diag_fn(match[i], L1 = L1, L2 = L2[i])))

# combine diagnostic summaries into one massive table
diags_table = rbind(
  parr_recruit_diags,
  pi_diags,
  phi_Pa_Mb_diags,
  phi_Mb_Ma_diags,
  phi_Ma_O0_diags,
  omega_diags,
  psi_O1_Rb_diags,
  psi_O2_Rb_diags,
  phi_O0_O1_diags,
  phi_O1_O2_diags,
  phi_Sb_Sa_diags,
  misc_ocean_surv_diags,
  z_diags,
  init_diags,
  stray_diags
)
```

This table displays summaries of MCMC diagnostics for all unobserved stochastic nodes in the model. Values in cells represent the percentage of parameters of a given type that fail a diagnostic check. E.g., 50% in the "Rhat > 1.05" column means that half of the parameters of that type had Rhat values greater than 1.05. Cells with no color indicate no offense and darker reds indicate greater frequency of an offense.

* **Rhat** measures convergence of multiple chains to a common target distribution. Values close to 1 are ideal, greater than 1.1 are problematic but possibly passable, and greater than 1.2 should not be used for inference (general rules).
* **ESS** measures the number of effective MCMC samples are available to estimate a given summary statistic of the posterior -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have fewer than 400 ESS -- which Vehtari et al. [2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full) suggest as a good rule of thumb.
* **MCCV** is Monte Carlo Standard Error of a summary statistic divided by the estimated posterior summary statistic, intended to represent a measure of precision for estimating posterior summary statistics -- here the posterior mean and quantiles of the 95% equal-tailed CRI. Percentages shown represent the fraction of parameters that have greater than 5% MCCV -- which Staton has (arbitrarily and at this time) chosen as the greatest amount of acceptable imprecision.

The calculations behind this table are implemented using the '[posterior](https://mc-stan.org/posterior/articles/posterior.html)' R package, which contains updated MCMC diagnostics (Vehtari et al. [2021](https://projecteuclid.org/journals/bayesian-analysis/advance-publication/Rank-Normalization-Folding-and-Localization--An-Improved-R%cb%86-for/10.1214/20-BA1221.full)).

```{r mcmc-diag-table}
# duplicate the diagnostic output
kable_table = diags_table

# order the rows
kable_table$L1 = factor(kable_table$L1, levels = unique(kable_table$L1))
kable_table = diags_table[order(kable_table$L1, kable_table$L2, kable_table$L3),]

# keep only relevant columns for kable
kable_table = kable_table[,-which(colnames(diags_table) %in% c("match", "mean_value", "Rhat_mean", "ess_bulk_mean"))]

# convert counts of offending parameters to percentages
kable_table$Rhat_1.05 = paste0(round(kable_table$Rhat_1.05/kable_table$n_params, 2) * 100, "%")
kable_table$Rhat_1.1 = paste0(round(kable_table$Rhat_1.1/kable_table$n_params, 2) * 100, "%")
kable_table$Rhat_1.2 = paste0(round(kable_table$Rhat_1.2/kable_table$n_params, 2) * 100, "%")
kable_table$ess_mean_400 = paste0(round(kable_table$ess_mean_400/kable_table$n_params, 2) * 100, "%")
kable_table$ess_q2.5_400 = paste0(round(kable_table$ess_q2.5_400/kable_table$n_params, 2) * 100, "%")
kable_table$ess_q97.5_400 = paste0(round(kable_table$ess_q97.5_400/kable_table$n_params, 2) * 100, "%")
kable_table$mccv_mean_0.05 = paste0(round(kable_table$mccv_mean_0.05/kable_table$n_params, 2) * 100, "%")
kable_table$mccv_q2.5_0.05 = paste0(round(kable_table$mccv_q2.5_0.05/kable_table$n_params, 2) * 100, "%")
kable_table$mccv_q97.5_0.05 = paste0(round(kable_table$mccv_q97.5_0.05/kable_table$n_params, 2) * 100, "%")

# add basic formatting
colnames(kable_table) = c("Model Component", "Parameter Type", "Group", "N", "> 1.05", "> 1.1", "> 1.2", "Mean", "q2.5", "q97.5", "Mean", "q2.5", "q97.5")
kable_table$Group[is.na(kable_table$Group)] = " "
kable_table$`Parameter Type`[is.na(kable_table$`Parameter Type`)] = " "

assign_color = function(x, colors = c("white","red")) {
  pal = colorRampPalette(colors)
  cols = pal(6)
  bin_names = levels(cut(rbeta(1e6, 1, 1) * 100, breaks = c(0,0.001,20,40,60,80,100), include.lowest = TRUE))
  names(cols) = bin_names
  bins = cut(x, breaks = c(0,0.001,20,40,60,80,100), include.lowest = TRUE)
  cols_use = cols[as.character(bins)]
  alpha(unname(cols_use), 0.5)
}

# produce the kable
kable(kable_table, row.names = FALSE, align = c("lllcccccccccc")) %>%
  kable_styling(full_width = TRUE, fixed_thead = TRUE, bootstrap_options = c("striped", "condensed")) %>%
  add_header_above(c(" " = 4, "Rhat" = 3, "ESS < 400" = 3, "MCCV > 5%" = 3)) %>%
  column_spec(2:3, width = "150px") %>%
  column_spec(c(4,7,10), border_right = TRUE) %>%
  column_spec(5, background = assign_color(as.numeric(stringr::str_remove(kable_table[,5], "%")))) %>%
  column_spec(6, background = assign_color(as.numeric(stringr::str_remove(kable_table[,6], "%")))) %>%
  column_spec(7, background = assign_color(as.numeric(stringr::str_remove(kable_table[,7], "%")))) %>%
  column_spec(8, background = assign_color(as.numeric(stringr::str_remove(kable_table[,8], "%")))) %>%
  column_spec(9, background = assign_color(as.numeric(stringr::str_remove(kable_table[,9], "%")))) %>%
  column_spec(10, background = assign_color(as.numeric(stringr::str_remove(kable_table[,10], "%")))) %>%
  column_spec(11, background = assign_color(as.numeric(stringr::str_remove(kable_table[,11], "%")))) %>%
  column_spec(12, background = assign_color(as.numeric(stringr::str_remove(kable_table[,12], "%")))) %>%
  column_spec(13, background = assign_color(as.numeric(stringr::str_remove(kable_table[,13], "%")))) %>%
  collapse_rows(columns = 1:2)
```
